Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-29 paper/paper.tex
Blair's changes

Solution-biased search allows for a much simpler parallel implementation. We
create a number of threads, and give each thread its own random seed, otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. At each restart each thread's new nogoods can be gathered, and
then each thread resumes, with a larger set of nogoods than in a sequential run. Finally, whenever
any single thread terminates, either due to having found a solution or proved unsatisfiability,
then every other thread may immediately terminate.

A number of factors combine to make this technique feasible:

\begin{itemize}
    \item Each thread will be run with the same restart schedule, and so will spend approximately
        the same amount of time between restarts. Because the only synchronisation between threads
        is at a restart, we expect threads to be busy doing search.  (This is in contrast
        to an alternative method for paralellising restarting search \cite{DBLP:conf/aaai/CireKS14},
        which packs together successive sequence values to produce a balanced workload.)
    \item Sequentially, on non-trivial instances the algorithm will restart often (many
        tens of thousands, for instances that reach the thousand second timeout).
    \item Because the search trees we explore are exponentially large, the randomness in the
        value-ordering heuristic is sufficient to ensure that most of the time, threads are
        exploring different parts of the search tree.
    \item The gathering of nogoods to describe the work done so far provides an alternative to
        requiring either a specific mechanism to allocate work, or expensive synchronisation between threads.
    \item If sometimes threads do happen to explore part of
        the same subproblems, this is not a problem: if the instance is satisfiable, either thread
        might find a solution first, and if the instance is unsatisfiable, we merely introduce some
        redundancy into the proof.\footnote{For solving a counting or enumeration problem,
        matters become slightly more complicated, but not devastatingly so.} The combination of
        rapid restarts and nogood recording is enough to ensure that this is only a small overhead.
\end{itemize}

@@

Because we are using nogood recording, an alternative approach is possible. Rather than using the
Luby sequence for restarts, we could restart after either a constant number of backtracks, or simply
after a certain time interval has passed. \Cref{figure:parallel} also shows the effects of restarting
every 100ms. Sequentially, this approach is slightly better than using the Luby sequence, being able to
solve 14,370 instances, with the last at 996.4 seconds. With thirty six threads, solving this many
instances takes 31.8 seconds, giving an aggregate speedup of 31.4.

@@

To further test the scalability of this technique, we also implemented a distributed-parallelism
layer on top of the threaded layer, using MPI. In contrast to the huge difficulties of implementing
work-stealing in a distributed memory setting, this required only the addition of a simple ``all
gather'' operation to communicate nogoods, and a ``gather'' to collect and combine the results of
each host. Termination, meanwhile, was handled by posting an empty nogood (and so termination can
only occur on a restart).

@@

    \caption{Above, the cumulative number of maximum common induced subgraph instances solved by
    k${\downarrow}$ and McSplit${\downarrow}$ over time, with and without restarts and nogood
    recording. Below, comparing k${\downarrow}$ (left) and McSplit${\downarrow}$ (right) on an instance by instance
    basis.}\label{figure:mcs}
\end{figure}

>>>>

Solution-biased search allows for a much simpler parallel implementation. We create a
number of threads, and give each thread its own random seed; otherwise each thread performs the same
sequential search. Threads synchronise on restarts, a simple barrier causing each thread to wait for
every other thread to also restart.  Nogoods from all threads are then gathered and combined before
search resumes, now with a larger set of nogoods than in a sequential run.  Finally, whenever any
single thread terminates, either due to having found a solution or proved unsatisfiability, then
every other thread may immediately terminate.

This technique requires only limited changes to the top level search driver, and none whatsoever to
the main recursive search algorithm. Notably, it does not require any locking or communication
during the recursive search, aside from a single atomic boolean flag to assist early termination.  A
number of factors combine to make this approach feasible:

\begin{itemize}
    \item Each thread will be run with the same restart schedule, and so will spend approximately
        the same amount of time between restarts. Because the only synchronisation between threads
        is at a restart, we expect threads to be busy doing search.  (This is in contrast
        to an alternative method for paralellising restarting search \cite{DBLP:conf/aaai/CireKS14},
        which packs together successive sequence values to produce a balanced workload.) This
        approach therefore avoids the irregular task size issues which usually arise in parallel
        combinatorial search.
    \item Sequentially, on non-trivial instances the algorithm will restart often (many
        tens of thousands, for instances that reach the thousand second timeout).
    \item Because the search trees we explore are exponentially large, the randomness in the
        value-ordering heuristic is sufficient to ensure that most of the time, threads are
        exploring different parts of the search tree.
    \item The gathering of nogoods to describe the work done so far provides an alternative to
        requiring either a specific mechanism to allocate work, or expensive synchronisation between threads.
    \item If sometimes threads do happen to explore part of
        the same subproblems, this is not a problem: if the instance is satisfiable, either thread
        might find a solution first, and if the instance is unsatisfiable, we merely introduce some
        redundancy into the proof.\footnote{For solving a counting or enumeration problem,
        matters become slightly more complicated, but not devastatingly so.} The combination of
        rapid restarts and nogood recording is enough to ensure that this is only a small overhead.
\end{itemize}

@@

Because we are using nogood recording, an alternative approach is possible. Rather than using the
Luby sequence for restarts, we could restart after either a constant number of backtracks, or simply
after a certain time interval has passed---this bounds the maximum possible idle time that threads
could spend at a barrier. \Cref{figure:parallel} also shows the effects of restarting every 100ms.
Sequentially, this approach is slightly better than using the Luby sequence, being able to solve
14,370 instances, with the last at 996.4 seconds. With thirty six threads, solving this many
instances takes 31.8 seconds, giving an aggregate speedup of 31.4.

@@

To further test the scalability of this technique, we also implemented a distributed-memory
parallelism layer on top of the threaded layer, using MPI. In contrast to the huge difficulties of
implementing work-stealing in a distributed memory setting, this required only the addition of two
MPI calls: an ``all gather'' operation to communicate nogoods, and a ``gather'' to collect and
combine the results of each host. Termination, meanwhile, was handled by posting an empty nogood
(and so termination can only occur on a restart).

@@

    \caption{Above, the cumulative number of maximum common induced subgraph instances solved by
    k${\downarrow}$ and McSplit${\downarrow}$ over time, with the two forms of search. Below,
    comparing k${\downarrow}$ (left) and McSplit${\downarrow}$ (right) on an instance by instance
    basis.}\label{figure:mcs}
\end{figure}
